---
title: "Gapminder"
author: "Reuben"
date: 2023-02-21
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)
```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
df_g = gapminder
head(df_g)
```

**Observations**:

-   The variables are:
    -   country

    -   continent

    -   year

    -   lifeExp

    -   pop

    -   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  df_g %>% 
    summarize(max(year))
year_min <- 
  df_g %>% 
    summarize(min(year))
year_max = as.integer(year_max)
year_min = as.integer(year_min)
# Last time you got mad because I used the dollar sign so I did it both ways.
#year_max <- max(gapminder$year)
#year_min <- min(gapminder$year)
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
df_g %>% 
  filter(year == min(year)) %>%
  ggplot(aes(gdpPercap, continent)) +
  geom_boxplot() +
  scale_x_log10()
```

**Observations**:

-   Oceania has a small range because there are so few countries
-   There is a big outlier in Asia, Kuwait.
-   Africa has the lowest average gdpPercap
-   Maybe less of an observation and just a thought but this plot makes me think about just how different gdpPercap and median gdp must be.

**Difficulties & Approaches**:

-   The main challenge was that the outliers made the graph hard to see. To fix this problem I took a log scale on the y axis.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
#Take an average of all the years
df_g_countries <-
  df_g %>% 
    group_by(country) %>%
    mutate(gdpPercap = mean(gdpPercap)) %>%
    filter(!duplicated(country))
#First method is to just take the values from the plot
df_g_countries$country[which(df_g_countries$gdpPercap %in% boxplot(df_g_countries$gdpPercap,plot=FALSE)$out)]

#Second method is to do all the calculations out and get the values that are 2 standard devations away from the mean
gdpPercap_c = df_g_countries$gdpPercap
mean = mean(gdpPercap_c)
std = sd(gdpPercap_c)
Tmin = mean-(2*std)
Tmax = mean+(2*std)
gdpPercap_outliers <- 
  df_g_countries %>% 
  filter(!between(gdpPercap, Tmin, Tmax))
gdpPercap_outliers
```

**Observations**:

-   Identify the outlier countries from q2
    -   Kuwait
    -   Norway
    -   Switzerland
    -   United States

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == max(year) | year == min(year)) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% 
      group_by(year) %>%
      filter(
          gdpPercap < quantile(gdpPercap, 0.01) | 
          gdpPercap > quantile(gdpPercap, 0.99)
      ),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  facet_wrap(~year)
```

**Observations**:

-   Kuwait was a big outlier and they are now less of one.
-   Other than Kuwait, the outliers have changed quite a bit.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
df_g_countries %>%
  ggplot(aes(lifeExp, gdpPercap)) +
  geom_point(aes(col=continent, size=pop)) +
  geom_smooth(method="loess", se=F) +
  labs(title="Life Expectancy Vs GDP Per Capita", y="GDP Per Capita", x="Life Expectancy")
```

-   Life expectancy and GDP per capita are correlated
-   European countries tend to have higher GDP per capitas then African countries.
-   It seems like higher life expediencies don't necessarily mean larger populations. Having information on birth rate might help us with figuring out what is going on with that.

```{r q5-task2}
## TASK: Your second graph
library(gganimate)
life_v_pop_plot <- 
  gapminder %>%
    ggplot(aes(lifeExp, pop, color = country)) +
    geom_point(alpha = 0.75, show.legend = FALSE) +
    scale_colour_manual(values = country_colors) +
    facet_wrap(~continent) +
    labs(title="Life Expectancy Vs Population", subtitle = "Year: {frame_time}", y="Population", x="Life Expectancy") +
    transition_time(year) +
    ease_aes("linear")
animate(life_v_pop_plot, duration = 12.5, fps = 20)
```

-   As time increases, both population and life expectancy tend to increase.
-   There is one interesting country in Africa that has its life expectancy drop a ton in 1991.
-   In Africa in particular, the life expectancy tends to spread out instead of all of the countries increasing equally.
-   Two countries in Asia are big outliers in terms of population growth.

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  filter(year == max(year) | year == min(year)) %>%
  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% 
      group_by(year) %>%
      filter(
          gdpPercap < quantile(gdpPercap, 0.01) | 
          gdpPercap > quantile(gdpPercap, 0.99)
      ),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  facet_wrap(~year)
```

-   The outliers in terms of life expectancy are disproportionately in Africa.
-   The range of life expectancy has shrunk slightly over time.
