---
title: "SAT and College Grades"
author: "Reuben Lewis"
date: 2020-04-10
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: How do we apply hypothesis testing to investigating data? In
this challenge you'll practice using hypothesis testing tools to make
sense of a dataset.

*Reading*: - [Harvard Study Says SATs Should Be Optional: Here's
Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why)
(Optional); easy-to-read news article on colleges going SAT-free -
[Norm-Referenced Tests and Race-Blind
Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university)
(Optional); technical report on relationship between the SAT/ACT and
non-academic factors

*Credit*: This is based on a [case
study](http://onlinestatbook.com/2/case_studies/sat.html) originally
prepared by Emily Zitek, with data collected through the research of
Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
library(ggdark)
```

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|--------------------------|-----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

*Background*: Every year about 2 million students take the Scholastic
Aptitude Test (SAT). The exam is
[controversial](http://www.nea.org/home/73288.htm) but [extremely
consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html).
There are many claims about the SAT, but we're going to look at just
one: Is the SAT predictive of scholastic performance in college? It
turns out this is a fairly complicated question to assess---we'll get an
introduction to some of the complexities.

# Obtain the Data

<!-- -------------------------------------------------- -->

### **q1** Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read_csv("./data/sat.csv")

## TODO: Do your "first checks"
head(df_sat)

df_sat %>% 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```

**Observations**:

-   Fill in the following "data dictionary"

| Column     | Meaning              |
|------------|----------------------|
| `high_GPA` | Highschool GPA       |
| `math_SAT` | Math Sat Score       |
| `verb_SAT` | Verbal Sat Score     |
| `comp_GPA` | Computer Science GPA |
| `univ_GPA` | University GPA       |

-   What information do we have about these students?
    -   Everything in the meaning section of the column.
-   What kinds of information *do we not have* about these students?
    -   We don't know their demographics (age, gender, etc.)
    -   We don't know their locations
    -   We don't know the dates the data was collected on
-   Based on these missing variables, what possible effects could be
    present in the data that we would have *no way of detecting*?
    -   Difficulty of course load

    -   Test taking skills

    -   Non graded activities

# Analysis with Hypothesis Testing

<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the
first based on hypothesis testing of correlation coefficients, and the
second based on fitting a regression model and interpreting the
regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)
```

## View 1: Correlations

<!-- ----------------------------------------------------------------------- -->

### **q2** Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO:
df_composite %>%
  pivot_longer(
    cols = c(high_GPA, both_SAT),
    names_to = "names",
    values_to = "values"
  ) %>% 
  ggplot(aes(x = values, y = univ_GPA)) +
  geom_point() +
  facet_wrap(~names, scales = "free") +
  dark_theme_minimal()
```

**Observations**:

-   What relationship do `univ_GPA` and `both_SAT` exhibit?
    -   Positive correlation
-   What relationship do `univ_GPA` and `high_GPA` exhibit?
    -   Positive correlation

### Hypothesis Testing with a Correlation Coefficient

<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation
coefficient. The idea is to set our null hypothesis to the case where
there is no correlation, and test to see if the data contradict that
perspective. Formally, the null (H0) and alternative (HA) hypotheses
relating to a correlation coefficient between two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the
assumption that `X, Y` are both normally distributed. First, let's check
to see if this assumption looks reasonable for our data.

### **q3** Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
df_composite %>% 
  ggplot() +
  geom_histogram(aes(x = both_SAT)) +
  dark_theme_minimal()
df_composite %>% 
  ggplot() +
  geom_histogram(aes(x = high_GPA)) +
  dark_theme_minimal()
df_composite %>% 
  ggplot() +
  geom_histogram(aes(x = univ_GPA)) +
  dark_theme_minimal()
```

**Observations**:

-   To what extent does `both_SAT` look like a normal distribution?
    -   Not very
-   To what extent does `high_GPA` look like a normal distribution?
    -   Not very
-   To what extent does `univ_GPA` look like a normal distribution?
    -   Not very

Keep in mind your findings as you complete q4.

### **q4** Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA
library(ggcorrplot)

corr <- round(cor(df_composite), 1)

ggcorrplot(corr, hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method="circle", 
           colors = c("orangered2", "white", "springgreen3"), 
           title="Correlogram", 
           ggtheme=dark_theme_minimal)
cor_test_result_1 <- cor.test(df_composite$high_GPA, df_composite$univ_GPA, method = "pearson")
cor_test_result_1
cor_test_result_2 <- cor.test(df_composite$both_SAT, df_composite$univ_GPA, method = "pearson")
cor_test_result_2
```

**Observations**:

-   Which correlations are significantly nonzero?
    -   All of them except verb_SAT and comp_GPA
-   Which of `high_GPA` and `both_SAT` seems to be more strongly
    correlated with `univ_GPA`?
    -   high_GPA
-   How do the results here compare with the visual you created in q2?
    -   There is a lot of correlation going on in both
-   Based on these results, what can we say about the predictive
    capabilities of both `high_GPA` and `both_SAT` to predict
    `univ_GPA`?
    -   high_GPA is a better predictor, but not by that much.

Finally, let's use the bootstrap to perform the same test using
*different* assumptions.

### **q5** Use the bootstrap to approximate a confidence interval.

Use the bootstrap to approximate a confidence interval for
`cor[high_GPA, univ_GPA]`. Compare your results---both the estimate and
confidence interval---to your results from q4.

*Hint 1*. The `cor(x, y)` function computes the correlation between two
variables `x` and `y`. You may find this more helpful than the
`cor.test()` function we used above.

*Hint 2*. You'll find that the documentation for `int_pctl` has some
**really** useful examples for this task!

```{r q5-task}
## TODO: Complete the following helper function to do a bootstrap analysis
corr_high_GPA <- function(split) {
  dat <- analysis(split)
  tibble(
    term = "cor",
    estimate = cor(dat$high_GPA, dat$univ_GPA, method = "spearman"),
  )
}

set.seed(69325)
## Use the bootstrap to approximate a CI
df_composite %>%
  bootstraps(1000) %>%
  mutate(estimates = map(splits, corr_high_GPA)) %>%
  int_pctl(estimates)
```

**Observations**:

-   How does your estimate from q5 compare with your estimate from q4?
    -   My estimate from q5 is 0.829909 and my estimate from q4 is
        0.7795631. They are pretty similar, but the one from q5 shows
        more correlation.
-   How does your CI from q5 compare with your CI from q4?
    -   My confidence interval from q5 is 0.7404223 to 0.8901223 and my
        confidence interval from q4 is 0.6911690 to 0.8449761. Both of
        them exclude the value 0, indicating that there is a
        statistically significant correlation. The range for q4 is
        slightly wider, but they are very close.

*Aside*: When you use two different approximations to compute the same
quantity and get similar results, that's an *encouraging sign*. Such an
outcome lends a bit more credibility to the results.

## View 2: Modeling

<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study
the relationship among more variables we can instead use a fitted model.
Using a model, we can also help assess whether it is *worthwhile* to
measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)
df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)
df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model

<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a
model, we can express our hypotheses in terms of the model parameters.
For instance, if we were interested in whether $X$ has an affect on $Y$,
we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on
$Y$. Let's apply this idea to relating the variables `univ_GPA` and
`high_GPA`. Luckily R has built-in tools to construct a confidence
interval on the $\beta$'s in a regression [1]; we'll simply use those
tools rather than do it by hand.

### **q6** Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
library(modelr)
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <- lm(univ_GPA ~ both_SAT, data = df_train)
## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
rsq_q6 <- modelr::rsquare(fit_basic, data = df_train)
rsq_q6
```

**Observations**:

-   What is the confidence interval on the coefficient of `both_SAT`? Is
    this coefficient significantly different from zero?
    -   The confidence interval is pretty close to zero.
-   By itself, how well does `both_SAT` predict `univ_GPA`?
    -   The r\^2 value being less than 0.4487683 indicates that there is
        a moderate fit.

Remember from `e-model03-interp-warnings` that there are challenges with
interpreting regression coefficients! Let's investigate that idea
further.

### **q7** Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
fit_basic <- lm(univ_GPA ~ high_GPA + both_SAT, data = df_train)
## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
rsq_q7 <- modelr::rsquare(fit_basic, data = df_train)
rsq_q7
```

**Observations**:

-   How well do these models perform, compared to the one you built in
    q6?
    -   This has a much larger confidence interval and a higher
        estimate.
-   What is the confidence interval on the coefficient of `both_SAT`
    when including `high_GPA` as a predictor?? Is this coefficient
    significantly different from zero?
    -   Its -0.0006740696 to 0.001741962 which is not significantly
        different from zero.
-   How do the hypothesis test results compare with the results in q6?
    -   The r\^2 value from q6 was 0.4487683 and the value from q7 was
        0.6061053. The higher value in q7 indicates a better fit.

## Synthesize

<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### **q8** Using the results from all previous q's, answer the following questions.

**Observations**:

-   Between `both_SAT` and `high_GPA`, which single variable would you
    choose to predict `univ_GPA`? Why?
    -   I would use high_GPA. high_GPA has a lower p-value compared to
        both_SAT.
-   Is `both_SAT` an effective predictor of `univ_GPA`? What specific
    pieces of evidence do you have in favor of `both_SAT` being
    effective? What specific pieces of evidence do you have against?
    -   It is an effective predictor in that the estimate is positive.
        On the other hand, the confidence interval is very small and
        includes zero, which means the effect could also be zero.
-   Based on the source of the data, how broadly do your findings apply?
    Do they apply to all students across the United States, or to a
    narrower subset of that population?
    -   There are only 105 people in the study, which is a small sample
        size. These results are only for a single university with a
        uniform grading system. All of the students are very similar
        already.

# End Notes

<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more
information see this [Wiki
article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
